Predicting _C. difficile_ Infection Severity from the Taxonomic Composition of the Gut Microbiome

### 2023-06-02 paper/paper.qmd
cite studies for feat importance

---
title: "Predicting _C. difficile_ Infection Severity from the Taxonomic Composition of the Gut Microbiome"
bibliography: references.bib
fontsize: 11pt
date: last-modified
date-format: medium
prefer-html: true
format:
  manuscript-pdf:
    mainfont: Helvetica
    keep-tex: true
    linenumbers: true
    doublespacing: true
    lof: false
  html:
    embed-resources: true
runningtitle: Predicting CDI Severity from OTUs
runningauthor: Sovacool
papertype: Preprint
author:
  - name: Kelly L. Sovacool
    orcid: 0000-0003-3283-829X
    affiliations:
      - ref: bioinf
  - name: Sarah E. Tomkovich
    affiliations:
      - ref: micro
  - name: Megan L. Coden
    affiliations:
      - ref: mcdb
  - name: Vincent B. Young
    affiliations:
      - ref: micro
      - ref: intmed
  - name: Krishna Rao
    affiliations:
      - ref: intmed
  - name: Patrick D. Schloss
    orcid: 0000-0002-6935-4275
    email: pschloss@umich.edu
    corresponding: true
    affiliations:
      - ref: micro
      - ref: ccmb
affiliations:
  - id: bioinf
    name: Department of Computational Medicine & Bioinformatics, University of Michigan
  - id: micro
    name: Department of Microbiology & Immunology, University of Michigan
  - id: mcdb
    name: Department of Molecular, Cellular, and Developmental Biology, University of Michigan
  - id: intmed
    name: Division of Infectious Diseases, Department of Internal Medicine, University of Michigan
  - id: ccmb
    name: Center for Computational Medicine and Bioinformatics, University of Michigan
#abstract: TODO
#importance: TODO
keywords: _C. difficile_ infection, supervised machine learning, gut microbiome, amplicion sequencing
---
```{r setup, eval=TRUE, echo=FALSE, cache=FALSE}
schtools::set_knitr_opts()
knitr::opts_chunk$set(dpi = 600)
```
```{r load_data, message = FALSE}
library(assertthat)
library(here)
library(kableExtra)
library(knitr)
library(tidyverse)

@@

prevalence of cdi. prevalence of severe cdi outcomes.

@@

We performed permutation feature importance to determine which OTUs contributed
the most to model performance (@fig-features).
An OTU was considered important if performance decreased when it was permuted in
at least 75% of the train/test splits.

@@

Discuss important OTUs. which ones concord with literature, which ones may be new.
Abundance data are sparse, likely due to these patients being on antibiotics.
Really showcases importance of having as many samples as possible when data are
sparse and the outcome is low prevalence.

@@

It's not enough for models to perform well to justify deploying them in a
clinical setting; benefit over current practices must be shown.
Estimating the NNB contextualizes model performance within clinical reality.
Amplicon sequencing is not typically performed for CDI patients, but if there is
clinical value to be gained by implementing OTU-based models, routinely
sequencing and profiling the microbial communities of CDI patients could be
justified.

@@

::: {#fig-performance fig-cap="**Performance of ML models.**
In the left facets, models were trained on the full datasets, with different
numbers of samples available for each severity definition.
In the right facets, models were trained on the same dataset consisting of the
intersection of samples with labels available for all definitions.
Note that the intersection dataset has exactly the same labels for attributable
and pragmatic severity, thus these have identical performance.
**A)** 
Area under the receiver-operator characteristic curve (AUROC) for the test sets
and cross-validation folds of the training sets, and the area under the balanced
precision-recall curve (AUBPRC) for the test sets.
Each point is annotated with the median performance across 100 train/test splits
with tails as the 95% CI.
**B)** Receiver-operator characteristic curves for the test sets.
Mean specificity is reported at each sensitivity value, 
with ribbons as the 95% CI.
**C)** Balanced precision-recall curves for the test sets.
Mean balanced precision is reported at each recall value, 
with ribbons as the 95% CI.
"}
![](figures/ml-performance.png)
:::

::: {#fig-features fig-cap="**Feature importance.**
**A)** Feature importance via permutation test.
For each OTU, the order of samples was randomized in the test set 100 times and
the performance was re-calculated to estimate the permutation performance.
An OTU was considered important if the performance decreased when the OTU was
permuted in at least 75% of the models.
OTUs with a greater difference in AUROC (actual performance minus permutation
performance) are more important.
Left: models were trained on the full datasets, with different numbers of
samples available for each severity definition.
Right: models were trained on the intersection of samples with all labels
available for each definition. Note that Attributable and Pragmatic severity are
exactly the same for the intersection dataset.
*Pseudomonas* (OTU 120) is not shown for IDSA severity in the full datasets nor
in the intersection dataset because it was removed during pre-processing due to
having near-zero variance.
**B)** Log~10~-transformed median relative abundances of the most important OTUs
on the full datasets, grouped by severity (shape). 
The vertical dashed line is the limit of detection.
"}
![](figures/feature-importance.png)
:::

>>>>

---
title: "Predicting _C. difficile_ Infection Severity from the Taxonomic Composition of the Gut Microbiome"
bibliography: references.bib
fontsize: 11pt
date: last-modified
date-format: medium
prefer-html: true
format:
  manuscript-pdf:
    keep-tex: true
    linenumbers: true
    doublespacing: true
    lof: false
  html:
    embed-resources: true
runningtitle: Predicting CDI Severity from OTUs
runningauthor: Sovacool
papertype: Preprint
author:
  - name: Kelly L. Sovacool
    orcid: 0000-0003-3283-829X
    affiliations:
      - ref: bioinf
  - name: Sarah E. Tomkovich
    affiliations:
      - ref: micro
  - name: Megan L. Coden
    affiliations:
      - ref: mcdb
  - name: Vincent B. Young
    affiliations:
      - ref: micro
      - ref: intmed
  - name: Krishna Rao
    affiliations:
      - ref: intmed
  - name: Patrick D. Schloss
    orcid: 0000-0002-6935-4275
    email: pschloss@umich.edu
    corresponding: true
    affiliations:
      - ref: micro
      - ref: ccmb
affiliations:
  - id: bioinf
    name: Department of Computational Medicine & Bioinformatics, University of Michigan
  - id: micro
    name: Department of Microbiology & Immunology, University of Michigan
  - id: mcdb
    name: Department of Molecular, Cellular, and Developmental Biology, University of Michigan
  - id: intmed
    name: Division of Infectious Diseases, Department of Internal Medicine, University of Michigan
  - id: ccmb
    name: Center for Computational Medicine and Bioinformatics, University of Michigan
#abstract: TODO
#importance: TODO
keywords: _C. difficile_ infection, supervised machine learning, gut microbiome, amplicion sequencing
---
```{r setup, eval=TRUE, echo=FALSE, cache=FALSE}
schtools::set_knitr_opts()
knitr::opts_chunk$set(dpi = 600)
```
```{r load_data, message = FALSE}
library(assertthat)
library(here)
library(kableExtra)
library(knitr)
library(tidyverse)

@@

prevalence of cdi. prevalence of severe cdi outcomes.
antibiotics typical risk factor for cdi, but nonantibiotic medications can 
increase susceptibilty too [@tomkovich_osmotic_2021]
and non-nosomial CDI on the rise

@@

We performed permutation feature importance to determine which OTUs contributed
the most to model performance.
An OTU was considered important if performance decreased when it was permuted in
at least 75% of the train/test splits, with greater differences in AUROC meaning
greater importance.
We plotted mean decrease in AUROC alongside log~10~-transformed mean
relative abundances for the top OTUs (@fig-features).
There is not always a clear pattern of increased or decreased relative abundance
for important OTUs in severe CDI cases, but all of the top 5 OTUs had an
increased mean relative abundance in severe cases relative to not severe cases.
_Enterococcus_ was the most important OTU, being significantly important for 
all models except for attributable severity on the full dataset.
Enrichment of _Enterococcus_ in _C. difficile_ infection and severity has been well-documented
in prior studies, thus its importance and increase in abundance for severe cases
is not suprising 
[@schubert_microbiome_2014;
@antharam_intestinal_2013;
@berkell_microbiota-based_2021;
@lesniak_gut_2022].
_Staphylococcus_ was important for the pragmatic and all-cause definitions on
the full datasets, but not for models trained on the intersection dataset.
_Lactobacillus_ was important only for the all-cause definition on the intersection
dataset, and its enrichment during _C. difficile_ infection has also been shown
in prior studies 
[@antharam_intestinal_2013;
@berkell_microbiota-based_2021].
All remaining OTUs had differences in AUROC < 0.02 and were only significantly
important in one or two of the models at most.
For many of the top OTUs, there is wide variance in importance, perhaps due to
the imbalanced nature of the severity outcomes.

@@

Discuss important OTUs. which ones concord with literature, which ones may be new.
Abundance data are sparse, likely due to these patients being on antibiotics.
Really showcases importance of having as many samples as possible when data are
sparse and the outcome is low prevalence.
we do not know which antibiotics were prescribed to treat these CDI cases, nor
which antibiotics patients may have taken prior to the CDI.
differences in microbiota between patients may be due to different abx...
different antibiotics have been shown to create different forms of dysbiotic
microbiota [@berkell_microbiota-based_2021] and
differential cdi clearance [@tomkovick_initial_2020].
vanc-resistant enterococcus new nosocomial alliance [@poduval_clostridium_2000].

@@

It's not enough for models to perform well to justify deploying them in a
clinical setting; benefit over current practices must be shown.
do no harm [@wiens_no_2019].
Estimating the NNB contextualizes model performance within clinical reality.
Amplicon sequencing is not typically performed for CDI patients, but if there is
clinical value to be gained by implementing OTU-based models, routinely
sequencing and profiling the microbial communities of CDI patients could be
justified.
resistance to vancomycin is increasing in staph, cdi, and enterococci -- even
more important to find alternate treatments.

@@

::: {#fig-performance fig-cap="**Performance of ML models.**
In the left facets, models were trained on the full datasets, with different
numbers of samples available for each severity definition.
In the right facets, models were trained on the same dataset consisting of the
intersection of samples with labels available for all definitions.
Note that the intersection dataset has exactly the same labels for attributable
and pragmatic severity, thus these have identical performance.
**A)** 
Area under the receiver-operator characteristic curve (AUROC) for the test sets
and cross-validation folds of the training sets, and the area under the balanced
precision-recall curve (AUBPRC) for the test sets.
Each point is annotated with the median performance across 100 train/test splits
with tails as the 95% CI.
**B)** Receiver-operator characteristic curves for the test sets.
Mean specificity is reported at each sensitivity value, 
with ribbons as the 95% CI.
**C)** Balanced precision-recall curves for the test sets.
Mean balanced precision is reported at each recall value, 
with ribbons as the 95% CI.
Original unbalanced precision-recall curves are shown in Figure S1.
"}
![](figures/ml-performance.png)
:::

::: {#fig-features fig-cap="**Feature importance.**
**A)** Feature importance via permutation test.
For each OTU, the order of samples was randomized in the test set 100 times and
the performance was re-calculated to estimate the permutation performance.
An OTU was considered important if the performance decreased when the OTU was
permuted in at least 75% of the models.
Mean difference in AUROC and the 75% confidence interval (CI) is reported for
each OTU, with starred OTUs being significant for the 75% CI.
OTUs with a greater difference in AUROC (actual performance minus permutation
performance) are more important.
Left: models were trained on the full datasets, with different numbers of
samples available for each severity definition.
Right: models were trained on the intersection of samples with all labels
available for each definition. Note that Attributable and Pragmatic severity are
exactly the same for the intersection dataset.
*Pseudomonas* (OTU 120) is not shown for IDSA severity in the full datasets nor
in the intersection dataset because it was removed during pre-processing due to
having near-zero variance.
**B)** Log~10~-transformed mean relative abundances of the most important OTUs
on the full datasets, grouped by severity (shape). 
The vertical dashed line is the limit of detection.
"}
![](figures/feature-importance.png)
:::
